---
title: "kolabas"
output: html_document
date: "2024-12-10"
---
Author: Mary-Ellen Lynall 2024

AIM: Test how similar the stress vs control meninges neutrophils and preneutrophils are to neutrophils from different locations (blood, various marrow samples)

Kolabas dataset:
Dataset downloaded from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE192616 on 11/12/24
Accompanying paper: https://pubmed.ncbi.nlm.nih.gov/37562402/ 

GEO: !Series_overall_design	"A single-cell RNA sequencing experiment was conducted using 10x Genomics technology on samples from three distinct groups, which included six bones, meninges, and the brain. The study comprised three groups: those that underwent MCAo surgery, those that underwent sham surgery, and a group of naive animals. Each sample consisted of cells pooled from three animals, with a total of 32 samples utilized. The MCAo surgery group contributed 2 samples, each consisting of cells pooled from 3 different animals. The sham and naive groups served as control groups. In addition to the single-cell RNA sequencing, bulk RNA sequencing experiments were conducted using the Illumina ligation stranded mRNA preparation kit. These experiments involved samples from the six bones under the three conditions (naive, sham, and MCAo). Separate measurements were taken for 4-6 animals per condition, yielding a total of 95 samples. Furthermore, bulk RNA sequencing was carried out using Smart-seq2 for mini-bulks of about 50 CD45+/Cd11b+ sorted immune cells each from the skull and femur. These samples were taken from wildtype animals (5 in total) and 5xFAD animals (3 in total), amassing 168 samples overall, with approximately 20 samples per animal."

!Sample_data_processing	"Supplementary_files_format_and_content: anndata (annotated data matrix) object saved in h5ad-file format (sc_adata_oct22m.h5ad, sc_adata_march22h.h5ad) as generated and used in the code at https://github.com/erturklab/skull_immune. The files contain the raw, preprocessed, normalized and annotated count data of all single cell experiments.

GPL21103 contains relevant data

-----
Also adding Evrard 2018 - bulk RNAseq
https://www.sciencedirect.com/science/article/pii/S1074761318300384?via%3Dihub

Aiming for the data underlying figure 3C
"To better understand the developmental relationship between preNeus, immature neutrophils (Ly6Glo/+ CXCR2−; termed immature Neus) and mature neutrophils (Ly6G+ CXCR2+; termed mature Neus), we sorted these three subsets and compared their morphology with sorted GMPs and blood neutrophils. Although GMP displayed a largely uncondensed nucleus with an immature cytosol, neutrophils progressively condensed their nucleus from a toroidal shape in preNeus to a poly-segmented shape in BM mature and blood neutrophils (Figure 3B). We also identified preNeus and immature Neus in the spleen (Figures S3B and S3E), but in much fewer numbers than in the BM (Figures S3C and S3D), and these cells were absent from the blood (data not shown).
We next used whole-transcriptome sequencing (RNA-seq) to determine how the molecular signatures of these neutrophil precursors differed. Principal-component analysis (PCA) of all transcripts revealed distinct gene expression profiles between all subsets (Figure 3C)."

Evrard count matrix at:
https://ftp.ncbi.nlm.nih.gov/geo/series/GSE109nnn/GSE109467/suppl/GSE109467%5FreadCount%5Frenamed%5FgeneName%2Etxt%2Egz

From Max Evrard (personal communication) 20250412

BN refers to blood neutrophils, with the numbers likely indicating different biological replicates and the ZT (Zeitgeber time) indicated by the number following 'BN' (e.g., BN13 would be at ZT13, and BN5 at ZT5).
EN refers to bone marrow immature neutrophils.
LN refers to bone marrow mature neutrophils.

```{r}

setwd("/Users/mary/non_dropbox/exps/exp042_meninges_stress_dropseq/")
library(here)
```

Now load Kolabas
```{r}

# Processed data are in supplementary so download
download_file_if_needed <- function(url, dest_path) {
  decompressed_file <- sub("\\.gz$", "", dest_path)  # Remove the ".gz" extension
  # Check if the unzipped file already exists
  if (!file.exists(decompressed_file)) {
    message("File not found at destination. Downloading...")
    # Download the file
    tryCatch({
      download.file(url, dest_path, mode = "wb") # Use mode "wb" for binary files
      R.utils::gunzip(dest_path, destname = decompressed_file, overwrite = TRUE)
      message("Download and unzip completed successfully.")
    }, error = function(e) {
      message("An error occurred during download: ", e$message)
    })
  } else {
    message("File already exists at destination. Skipping download.")
  }
}


# Big file
url <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE192nnn/GSE192616/suppl/GSE192616%5Fsc%5Fadata%5Foct22m%2Eh5ad%2Egz"
dest_path <- "/Users/mary/Documents/research/exps/public/scrnaseq/kolabas2023/GSE192616_sc_adata_oct22m.h5ad.gz" # 1194 MB
download_file_if_needed(url, dest_path)

# Now read the decompressed file
# BiocManager::install("zellkonverter")
library(zellkonverter)
adata <- zellkonverter::readH5AD(sub("\\.gz$", "", dest_path)) 

```

View
```{r}

library(tidyverse)
library(SingleCellExperiment)
library(tidySingleCellExperiment)
library(magrittr)

# raw_counts and X (log norm counts)
assays(adata)

# Check info. 32 samples, 8 tissues
adata
conditions <- adata %>% colData() %>% as_tibble() %>% select(sample_id, region, condition, cell_types) %>% unique()
conditions$cell_types %>% unique() # There are "pro  neutrophils" and "neutrophils" here
conditions$region %>% unique() # Brain Femur Humerus Meninges Pelvis Scapula Skull Vertebra

# Drop everyting except neuts 
# adata <- adata %>% filter(cell_types %in% c("neutrophils"))

# Drop everything except neuts and preneutrophils
adata <- adata %>% filter(cell_types %in% c("pro neutrophils","neutrophils"))

# Make naming consistent with other datasets
adata$cell_types %<>% as.character()
adata$cell_types[adata$cell_types %in% "pro neutrophils"] <- "Preneutrophils"
adata$cell_types[adata$cell_types %in% "neutrophils"] <- "Neutrophils"

tmp <- adata %>% colData() %>% as_tibble() %>% select(sample_id, region, condition, cell_types) %>% unique() 
# View all conditions (for which we have neuts)
# One sham, one naive and two MCAO for each condition
tmp %>% arrange(region) %>% as.data.frame()

```

```{r}

# get a summarized experiment
adata$cellgroup <- paste(adata$sample_id, adata$region, adata$condition, adata$cell_types, sep="_")

# These only have one neutrophil cell so drop otherwise aggregate_cells can't run
adata[,adata$cellgroup %in% c("MUC12842_Brain_MCAO_neutrophils", "MUC12850_Brain_Naive_neutrophils")]

if ("cellgroup" %in% colnames(colData(adata))) {
  adata <- adata[, !(adata$cellgroup %in% c("MUC12842_Brain_MCAO_neutrophils","MUC12850_Brain_Naive_neutrophils"))]
} else {
  stop("The column 'cellgroup' does not exist in colData(adata).")
}

summary(as.vector(assay(adata, "raw_counts")[1:100, 1:10]))
summary(as.vector(assay(adata, "X")[1:100, 1:10]))
lib_sizes <- colSums(assay(adata))
summary(lib_sizes)

# Drop non-needed data as interferes with joins
# Pseudobulk based on RAW counts
minimal_sce <- SingleCellExperiment::SingleCellExperiment(
  assays = list(raw_counts = assay(adata, "raw_counts")),
  colData = colData(adata)
)

pseudo_bulk <-
  aggregate_cells(minimal_sce, c(sample_id, region, condition, cell_types))

# Check low cell numbers
pseudo_bulk %>% colData %>% as_tibble() %>% arrange(.aggregated_cells)

# Remove samples with <25 aggregated cells
library(tidySummarizedExperiment)
pseudo_bulk_kolabas <- pseudo_bulk %>% filter(.aggregated_cells>=25)

pseudo_bulk_kolabas %>% as_tibble() %>% arrange(.aggregated_cells) 

# Print cells per condition
cell_counts_df <- colData(pseudo_bulk_kolabas) %>%
  as_tibble() %>%
  mutate(sample = colnames(pseudo_bulk_kolabas)) %>%
  select(sample, .aggregated_cells) %>%
  arrange(desc(.aggregated_cells))

cell_counts_df %>% as.data.frame()

# Drop stroke and also sham, as sham had big effect on transcriptome. Keep only naive cells for building PCA.
pseudo_bulk_kolabas <- pseudo_bulk_kolabas %>% filter(condition %in% c("Naive"))

```


VST the Kolabas
```{r}
library(DESeq2)

# Plot library sizes for Kolabas pseudobulk
library_sizes <- colSums(assay(pseudo_bulk_kolabas, "raw_counts"))
libsize_df <- tibble(sample = names(library_sizes), libsize = library_sizes)

ggplot(libsize_df, aes(x = reorder(sample, -libsize), y = libsize)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Library sizes of Kolabas pseudobulk samples", x = "Sample", y = "Total raw counts") +
  theme_minimal()

# Pivot pseudobulk counts to a gene x sample matrix
kolabas_counts <- pseudo_bulk_kolabas %>%
  as_tibble() %>%
  select(.sample, .feature, raw_counts) %>%
  pivot_wider(names_from = .sample, values_from = raw_counts) %>%
  column_to_rownames(var = ".feature") %>%
  as.matrix()

# Step 2: Create colData for DESeq2
sample_meta <- colData(pseudo_bulk_kolabas)[colnames(kolabas_counts), , drop = FALSE]

# Construct DESeqDataSet
dds_kolabas <- DESeqDataSetFromMatrix(
  countData = round(kolabas_counts),
  colData = sample_meta,
  design = ~ 1
)

# Filter out low-count genes 
dds_kolabas <- dds_kolabas[rowSums(counts(dds_kolabas)) > 10, ]

dds_libsize <- colSums(counts(dds_kolabas))
dds_libsize_df <- tibble(sample = names(dds_libsize), libsize = dds_libsize)

# Apply VST
vst_kolabas <- vst(dds_kolabas, blind = TRUE)
vst_mat_kolabas <- assay(vst_kolabas)

# Look at post-VST values
vst_long <- reshape2::melt(vst_mat_kolabas)
colnames(vst_long) <- c("Gene", "Sample", "Expression")

ggplot(vst_long, aes(x = Sample, y = Expression)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(title = "VST Expression Distributions per Sample", y = "VST Expression", x = "Sample") +
  coord_flip() +
  theme_minimal()

# Add VST back to pseudo_bulk_kolabas
common_genes <- intersect(rownames(vst_mat_kolabas), rownames(pseudo_bulk_kolabas))
vst_mat_kolabas <- vst_mat_kolabas[common_genes, , drop = FALSE]
pseudo_bulk_kolabas <- pseudo_bulk_kolabas[common_genes, ]
assay(pseudo_bulk_kolabas, "vst") <- vst_mat_kolabas
```

PCA all Kolabas samples
```{r}

pca <- prcomp(t(assay(pseudo_bulk_kolabas, "vst")), scale. = TRUE)
pca_df <- as.data.frame(pca$x)
pca_df$sample <- colnames(pseudo_bulk_kolabas)

ggplot(pca_df, aes(x = PC1, y = PC2, label = sample)) +
  geom_point() +
  geom_text(size = 2, vjust = -1) +
  theme_minimal() +
  labs(title = "PCA of Pseudobulk VST", x = "PC1", y = "PC2")

# Drop severe outliers
samples_to_remove <- c("MUC12846___Meninges___Naive___Preneutrophils","MUC12846___Meninges___Naive___Neutrophils")

pseudo_bulk_kolabas <- pseudo_bulk_kolabas[, !colnames(pseudo_bulk_kolabas) %in% samples_to_remove]

```


Load Kigar stressed meninges neutrophils and preneutrophils
```{r}
# Load sce object
combi <- readRDS(paste0(here::here("res/"),"combi_clustered_leiden_annotated.RDS"))

load(file=here::here("res/amb_to_drop.R"))
sce <- combi[!rowData(combi)$ID %in% amb_to_drop,]
sce <- sce[,!sce$Annotation == "Erythrocytes"]

neuts <- sce[,sce$Annotation %in% c("Preneutrophils","Neutrophils")]
```


Make Kigar pseudobulks from raw counts and apply VST
```{r}

# Make pseudobulk based on raw counts
pseudo_bulk_stress <- neuts %>%
  tidySingleCellExperiment::aggregate_cells(c(group, cond, Annotation), assays = "counts")

# Check cell numbers
pseudo_bulk_stress %>% colData %>% as_tibble() %>% arrange(.aggregated_cells) # Minimum 25 like Kolabas

# Plot library sizes for stress pseudobulks
library_sizes <- colSums(assay(pseudo_bulk_stress, "counts"))
libsize_df <- tibble(sample = names(library_sizes), libsize = library_sizes)

ggplot(libsize_df, aes(x = reorder(sample, -libsize), y = libsize)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Library sizes of Stress pseudobulk samples", x = "Sample", y = "Total raw counts") +
  theme_minimal()

# Pivot pseudobulk counts to a gene x sample matrix
stress_counts <- pseudo_bulk_stress %>%
  as_tibble() %>%
  select(.sample, .feature, counts) %>%
  pivot_wider(names_from = .sample, values_from = counts) %>%
  column_to_rownames(var = ".feature") %>%
  as.matrix()

# Create colData for DESeq2
sample_meta <- colData(pseudo_bulk_stress)[colnames(stress_counts), , drop = FALSE]

# Construct DESeqDataSet
dds_stress <- DESeqDataSetFromMatrix(
  countData = round(stress_counts),
  colData = sample_meta,
  design = ~ 1
)

# Filter out low-count genes - NO, keep all genes, as filtering later by hvg intersect
# dds_stress <- dds_stress[rowSums(counts(dds_stress)) > 10, ]

# Apply VST
vst_stress <- vst(dds_stress, blind = TRUE)
vst_mat_stress <- assay(vst_stress)

# Look at post-VST values
vst_long <- reshape2::melt(vst_mat_stress)
colnames(vst_long) <- c("Gene", "Sample", "Expression")

ggplot(vst_long, aes(x = Sample, y = Expression)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(title = "VST Expression Distributions per Sample", y = "VST Expression", x = "Sample") +
  coord_flip() +
  theme_minimal()

# Add VST back to pseudo_bulk_stress
common_genes <- intersect(rownames(vst_mat_stress), rownames(pseudo_bulk_stress))
vst_mat_stress <- vst_mat_stress[common_genes, , drop = FALSE]
pseudo_bulk_stress <- pseudo_bulk_stress[common_genes, ]
assay(pseudo_bulk_stress, "vst") <- vst_mat_stress
```


Download Evrard 2018 data (GEO supplementary file)
```{r}

matrix_file = '/Users/mary/Documents/research/exps/public/bulkrnaseq/evrard2018/GSE109467_readCount_renamed_geneName.txt'

# Read the data
raw_counts <- read_tsv(matrix_file, col_types = cols(.default = "c"))

colnames(raw_counts)[1] <- "gene_id"

# Convert expression columns to numeric
expr_numeric <- raw_counts %>%
  select(-Chr, -Start, -End, -Strand, -Length) %>%
  mutate(across(-c(1, 2), as.numeric))  # keep gene_id and gene_name as character

# Add average expression
expr_filtered <- expr_numeric %>%
  rowwise() %>%
  mutate(avg_expr = mean(c_across(-c(gene_id, gene_name)), na.rm = TRUE)) %>%
  ungroup()

# Keep the row with highest avg expression for each gene_name
expr_best <- expr_filtered %>%
  group_by(gene_name) %>%
  slice_max(order_by = avg_expr, n = 1, with_ties = FALSE) %>%
  ungroup()

# Now make it a matrix with gene_name as rownames
expr_counts <- expr_best %>%
  select(-gene_id, -avg_expr) %>%
  column_to_rownames("gene_name") %>%
  as.matrix()

names <- colnames(expr_counts)

colnames(expr_counts) <- paste0(names, "___", ifelse(startsWith(names, "BN"), "Neutrophils___Blood",
                         ifelse(startsWith(names, "EN"), "Immature neutrophils___Femur",
                         ifelse(startsWith(names, "LN"), "Neutrophils___Femur",
                         ifelse(startsWith(names, "PN"), "Preneutrophils___Femur",
                         ifelse(startsWith(names, "GMP"), "GMP___Femur", names))))))                               
               
# Create DESeq2 dataset
dds <- DESeqDataSetFromMatrix(countData = expr_counts,
                              colData = data.frame(row.names = colnames(expr_counts)),
                              design = ~ 1)

# Filter genes with very low counts
# dds <- dds[rowSums(counts(dds)) >= 10, ]

# Normalize with variance stabilizing transformation
vsd <- vst(dds, blind = TRUE)

# Extract normalized matrix
vsd_mat <- assay(vsd)

# Convert to long format for ggplot
library(reshape2)
vsd_df <- melt(vsd_mat)
colnames(vsd_df) <- c("Gene", "Sample", "Expression")

# Boxplot
ggplot(vsd_df, aes(x = Sample, y = Expression)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(title = "Normalized Expression Distributions (VST)", y = "Expression", x = "Sample") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Create a SummarizedExperiment
bulk_evrard <- SummarizedExperiment::SummarizedExperiment(
  assays = list(vst = vsd_mat),
  colData = colData(vsd),          # Use sample metadata from DESeq2 object
  rowData = rowData(vsd)           # Keep gene metadata if available
)

# Drop GMP 
keep <- !startsWith(names, "GMP") 

# Subset the SummarizedExperiment
bulk_evrard <- bulk_evrard[, keep]

```

Make Kolabas pseudo_bulk and calculate hvgs.
```{r}
library(tidyr)
library(dplyr)
library(magrittr)

pseudo_bulk_kolabas$condition %<>% droplevels()

dim(pseudo_bulk_kolabas) 

tmp <- pseudo_bulk_kolabas %>%
  as_tibble() %>%
  select(.sample, .feature, vst) %>%
  pivot_wider(names_from = .feature, values_from = vst)

# Calculate the variance for each gene (column)
gene_variances <- apply(tmp[,-1], 2, var)

# Take top 20% most variable 
threshold <- quantile(gene_variances, 0.8, na.rm=T)

# Filter for highly variable genes
highly_variable_genes <- names(gene_variances[gene_variances >= threshold])
```

Get hvg common to all 3 datasets
```{r}

# Get common HVGs across all three datasets
highly_variable_genes_intersect <- highly_variable_genes[
  highly_variable_genes %in% rownames(pseudo_bulk_stress) &
  highly_variable_genes %in% rownames(bulk_evrard)
]

print(length(highly_variable_genes)) # 2287
print(length(highly_variable_genes_intersect)) # 1999

```


Make Kolabas PCA with HVG intersect
```{r}

# Reorder Kolabas to only HVG intersect
pseudo_bulk_kolabas <- pseudo_bulk_kolabas[highly_variable_genes_intersect, ]

pca_matrix_kolabas <- pseudo_bulk_kolabas %>%
  as_tibble() %>%
  select(.sample, .feature, vst) %>%
  pivot_wider(names_from = .feature, values_from = vst) %>% select(all_of(c(".sample", highly_variable_genes_intersect)))

# Remove the `.sample` column and convert to matrix
pca_matrix_kolabas_values <- pca_matrix_kolabas %>%
  select(-.sample) %>%
  as.matrix()

# Check rownames
rownames(pca_matrix_kolabas_values) <- pca_matrix_kolabas$.sample

# Perform PCA
pca_result <- prcomp(pca_matrix_kolabas_values, center = T, scale. = TRUE)

# View the PCA result
summary(pca_result)  # Check explained variance
# pca_result$x %>% head        # Principal component scores for each sample
# PC1 is 93% variance, PC2 is 1.9%, PC3 is 1.1%, PC4 is 1.0%

```

Subset Kigar to hvg intersect and apply Kolabas PCA transformation
```{r}

# Subset stress to hvg intersect
pseudo_bulk_stress_hvg <- pseudo_bulk_stress %>%
  filter(.feature %in% highly_variable_genes_intersect) 

# Ensure reorder rows for both
pseudo_bulk_stress_hvg <- pseudo_bulk_stress_hvg[ highly_variable_genes_intersect, ]

# Check TRUE
identical(rownames(pseudo_bulk_kolabas), rownames(pseudo_bulk_stress_hvg))

pca_matrix_stress <- pseudo_bulk_stress_hvg %>%
  as_tibble() %>%
  select(.sample, .feature, vst) %>%
  pivot_wider(names_from = .feature, values_from = "vst")

# Remove the `.sample` column and convert to matrix
pca_matrix_stress_values <- pca_matrix_stress %>%
  select(-.sample) %>%
  as.matrix()

# Check rownames
rownames(pca_matrix_stress_values) <- pca_matrix_stress$.sample

# Quantile normalize the matrix to better match distribution in Kolabas
library(preprocessCore)

# Make sure both matrices are numeric and have no NAs
stopifnot(is.numeric(pca_matrix_stress_values), !any(is.na(pca_matrix_stress_values)))
stopifnot(is.numeric(pca_matrix_kolabas_values), !any(is.na(pca_matrix_kolabas_values)))

# Quantile normalize stress values to match the distribution of pca_matrix_kolabas_values
pca_matrix_stress_qn <- t(normalize.quantiles.use.target(
  t(as.matrix(pca_matrix_stress_values)),
  target = sort(colMeans(pca_matrix_kolabas_values))
))

# Restore row and column names
rownames(pca_matrix_stress_qn) <- rownames(pca_matrix_stress_values)
colnames(pca_matrix_stress_qn) <- colnames(pca_matrix_stress_values)

# Scale the new sample using the PCA model parameters
new_samples_scaled <- scale(pca_matrix_stress_qn, center = pca_result$center, scale = pca_result$scale)

# Project the scaled sample onto the PCA components by matrix multiplication
new_sample_pca <- new_samples_scaled %*% pca_result$rotation

# Add labels
kolabas_pca <- as.data.frame(pca_result$x)
kolabas_pca$sample <- rownames(kolabas_pca)
kolabas_pca$region <- str_split_i(rownames(kolabas_pca),"___",2)
kolabas_pca$dataset <- "Kolabas"
kolabas_pca$condition <- str_split_i(rownames(kolabas_pca),"___",3)
kolabas_pca$dataset <- "Kolabas"
kolabas_pca$celltype <- str_split_i(rownames(kolabas_pca),"___",4)

# Match naming to stress dataset
kolabas_pca$condition[kolabas_pca$condition %in% "Naive"] <- "HC"
kolabas_pca %>% sample_n(6)

# Add stress dataset labels
new_sample_pca <- as.data.frame(new_sample_pca)
new_sample_pca$sample <- rownames(new_sample_pca)
new_sample_pca$region <- "Meninges"
new_sample_pca$condition <- str_split_i(rownames(new_sample_pca),"___",2)
new_sample_pca$dataset <- "Kigar"
new_sample_pca$celltype <- str_split_i(rownames(new_sample_pca),"___",3)
new_sample_pca %>% sample_n(6)

# Bind original PCA scores with the new sample's PCA scores
pca_scores_combined <- as.data.frame(rbind(kolabas_pca, new_sample_pca))

library(RColorBrewer)
# PC1 represents cell type
ggplot(pca_scores_combined, aes(x=PC1, y=PC2)) + geom_point(aes(colour=celltype), size=3) + scale_color_brewer(palette = "Paired") + theme_light()

# PC2 represents tissue of origion
ggplot(pca_scores_combined, aes(x=PC1, y=PC2)) + geom_point(aes(colour=region), size=3) + scale_color_brewer(palette = "Paired") + theme_light()

library(ggrepel)
ggplot(pca_scores_combined, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = region), size = 3) +
  geom_text_repel(aes(label = sample), size = 4, max.overlaps = 100) +
  scale_color_brewer(palette = "Paired") +
  theme_light() +
  labs(title = "PCA with Sample Labels", x = "PC1", y = "PC2")
#ggsave(here("pics/kolabas_neutrophils_PC2.pdf"), width=4,height=4)

# PC3 reflects dataset
ggplot(pca_scores_combined, aes(x=PC3, y=PC4)) + geom_point(aes(colour=dataset), size=4) + scale_color_brewer(palette = "Paired") + theme_light()

ggplot(pca_scores_combined, aes(x=PC4, y=PC3)) + geom_point(aes(colour=condition), size=4) + scale_color_brewer(palette = "Paired") + theme_light()

```


Now add the Evrard PCA
```{r}

# Subset evrard to hvg intersect
bulk_evrard_hvg <- bulk_evrard %>%
  filter(.feature %in% highly_variable_genes_intersect) 

# Ensure reorder rows to match stress and kolabas above
bulk_evrard_hvg <- bulk_evrard_hvg[ highly_variable_genes_intersect, ]

# Check TRUE
identical(rownames(pseudo_bulk_kolabas), rownames(bulk_evrard_hvg))

pca_matrix_evrard <- bulk_evrard_hvg %>%
  as_tibble() %>%
  select(.sample, .feature, vst) %>%
  pivot_wider(names_from = .feature, values_from = "vst")

# Remove the `.sample` column and convert to matrix
pca_matrix_evrard_values <- pca_matrix_evrard %>%
  select(-.sample) %>%
  as.matrix()

# Check rownames
rownames(pca_matrix_evrard_values) <- pca_matrix_evrard$.sample

# Make sure both matrices are numeric and have no NAs
stopifnot(is.numeric(pca_matrix_evrard_values), !any(is.na(pca_matrix_evrard_values)))
stopifnot(is.numeric(pca_matrix_kolabas_values), !any(is.na(pca_matrix_kolabas_values)))

# Quantile normalize evrard values to match the distribution of pca_matrix_kolabas_values
pca_matrix_evrard_qn <- t(normalize.quantiles.use.target(
  t(as.matrix(pca_matrix_evrard_values)),
  target = sort(colMeans(pca_matrix_kolabas_values))
))

# Restore row and column names
rownames(pca_matrix_evrard_qn) <- rownames(pca_matrix_evrard_values)
colnames(pca_matrix_evrard_qn) <- colnames(pca_matrix_evrard_values)

# Scale the new sample using the PCA model parameters
new_samples_scaled_evrard <- scale(pca_matrix_evrard_qn, center = pca_result$center, scale = pca_result$scale)

# Project the scaled sample onto the PCA components by matrix multiplication
evrard_pca <- as.data.frame(new_samples_scaled_evrard %*% pca_result$rotation)

evrard_pca$sample <- str_split_i(rownames(evrard_pca),"___",1)
evrard_pca$region <- str_split_i(rownames(evrard_pca),"___",3)
evrard_pca$condition <- "HC"
evrard_pca$dataset <- "Evrard"
evrard_pca$celltype <- str_split_i(rownames(evrard_pca),"___",2)
evrard_pca %>% sample_n(6)

# Combine original PCA scores with the new sample's PCA scores
pca_scores_combined_3 <- as.data.frame(rbind(pca_scores_combined, evrard_pca))

ggplot(pca_scores_combined_3, aes(x=PC1, y=PC2)) + geom_point(aes(colour=region), size=2) + scale_color_brewer(palette = "Paired") + theme_light()

```

Final plots
```{r}

# Region plot
ggplot(pca_scores_combined_3, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = region, shape = dataset), size = 3) +
  #geom_text_repel(aes(label = sample), size = 3, max.overlaps = 100) +
  scale_color_brewer(palette = "Paired") +
  theme_light() +
  labs(title = "", x = "PC1", y = "PC2")


ggplot(pca_scores_combined_3, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = celltype, shape = dataset), size = 3) +
  #geom_text_repel(aes(label = sample), size = 3, max.overlaps = 100) +
  scale_color_brewer(palette = "Paired") +
  theme_light() +
  labs(title = "", x = "PC1", y = "PC2")

ggplot(pca_scores_combined_3, aes(x = PC1, y = PC2)) +
  geom_point(aes(shape = dataset, fill = celltype, colour = condition), size = 3, stroke = 0.4) +
  scale_shape_manual(values = c(21, 22, 23, 24, 25)) +
  scale_fill_brewer(palette = "Paired") +
  scale_colour_manual(values = c("white", "black")) +
  guides(
    fill = guide_legend(override.aes = list(shape = 22, colour = "black")), # Show fill colors with black border
    colour = guide_legend(override.aes = list(shape = 22, fill = "white"))  # Show border color only
  ) +
  labs(title = "", x = "PC1", y = "PC2") +
  theme_minimal()

ggsave(here("pics/neutrophils_3datasets_condition_celltype.pdf"), width=5, height=5)

ggplot(pca_scores_combined_3, aes(x = PC1, y = PC2)) +
  geom_point(aes(shape = dataset, fill = region, colour = condition), size = 3, stroke = 0.4) +
  scale_shape_manual(values = c(21, 22, 23, 24, 25)) +
  scale_fill_brewer(palette = "Paired") +
  scale_colour_manual(values = c("white", "black")) +
  guides(
    fill = guide_legend(override.aes = list(shape = 22, colour = "black")), # Show fill colors with black border
    colour = guide_legend(override.aes = list(shape = 22, fill = "white"))  # Show border color only
  ) +
  labs(title = "", x = "PC1", y = "PC2") +
  theme_minimal()

ggsave(here("pics/neutrophils_3datasets_condition_region.pdf"), width=5, height=5)

ggplot(pca_scores_combined_3, aes(x=PC2, y=PC3)) + geom_point(aes(colour=region), size=2) + scale_color_brewer(palette = "Paired") + theme_light()
ggsave(here("pics/three_datasets_PC2_PC3.pdf"), width=5,height=5)

```


```{r}

sessionInfo()

```
