---
title: "kolabas"
output: html_document
date: "2024-12-10"
---
Author: Mary-Ellen Lynall 2024

Dataset downloaded from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE192616 on 11/12/24

Accompanying paper: https://pubmed.ncbi.nlm.nih.gov/37562402/ 

In Kolabas 2023 there are scRNAseq data from meninges, brain, calvaria, other bones. 

AIM: Test how similar the meninges neutrophils are to neutrophils from each of those locations.
HYPOTHESIS: Our stress-associated neutrophils are most similar to Kolabas meninges and closer to Kolabas calvaria than to other bones in the Kolabas dataset.

```{r}

setwd("/Users/mary/non_dropbox/exps/exp042_meninges_stress_dropseq/")
library(here)
```


```{r}

# Processed data are in supplementary so download

download_file_if_needed <- function(url, dest_path) {
  decompressed_file <- sub("\\.gz$", "", dest_path)  # Remove the ".gz" extension
  # Check if the unzipped file already exists
  if (!file.exists(decompressed_file)) {
    message("File not found at destination. Downloading...")
    # Download the file
    tryCatch({
      download.file(url, dest_path, mode = "wb") # Use mode "wb" for binary files
      R.utils::gunzip(dest_path, destname = decompressed_file, overwrite = TRUE)
      message("Download and unzip completed successfully.")
    }, error = function(e) {
      message("An error occurred during download: ", e$message)
    })
  } else {
    message("File already exists at destination. Skipping download.")
  }
}


# Big file
url <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE192nnn/GSE192616/suppl/GSE192616%5Fsc%5Fadata%5Foct22m%2Eh5ad%2Egz"
dest_path <- "/Users/mary/Documents/research/exps/public/scrnaseq/kolabas2023/GSE192616_sc_adata_oct22m.h5ad.gz" # 1194 MB
download_file_if_needed(url, dest_path)

# Now read the decompressed file
library(zellkonverter)
adata <- zellkonverter::readH5AD(sub("\\.gz$", "", dest_path))

```

View
```{r}

library(tidyverse)
library(SingleCellExperiment)
library(tidySingleCellExperiment)

# Drop raw data as very large
assay(adata, "raw_counts") <- NULL

# Check info. 32 samples, 8 tissues
adata
conditions <- adata %>% colData() %>% as_tibble() %>% select(sample_id, region, condition, cell_types) %>% unique()
conditions$cell_types %>% unique() # There are "pro neutrophils" and "neutrophils" here
conditions$region %>% unique() # Brain Femur Humerus Meninges Pelvis Scapula Skull Vertebra

# Drop everyting except neuts 
adata <- adata %>% filter(cell_types %in% c("neutrophils"))

tmp <- adata %>% colData() %>% as_tibble() %>% select(sample_id, region, condition) %>% unique() 
# View all conditions (for which we have neuts)
# One sham, one naive and two MCAO for each condition
tmp %>% arrange(region) %>% as.data.frame()

```

```{r}
# get a summarized experiment
adata$cellgroup <- paste(adata$sample_id, adata$region, adata$condition, adata$cell_types, sep="_")

# These only have one neutrophil cell so drop otherwise aggregate_cells can't run
adata[,adata$cellgroup %in% c("MUC12842_Brain_MCAO_neutrophils", "MUC12850_Brain_Naive_neutrophils")]

if ("cellgroup" %in% colnames(colData(adata))) {
  adata <- adata[, !(adata$cellgroup %in% c("MUC12842_Brain_MCAO_neutrophils","MUC12850_Brain_Naive_neutrophils"))]
} else {
  stop("The column 'cellgroup' does not exist in colData(adata).")
}

pseudo_bulk <- adata %>%
  tidySingleCellExperiment::aggregate_cells(c(sample_id, region, condition, cell_types))

# Check low cell numbers
pseudo_bulk %>% colData %>% as_tibble() %>% arrange(.aggregated_cells)

# Remove samples with <10 aggregated cells
library(tidySummarizedExperiment)
pseudo_bulk_filtered <- pseudo_bulk %>% filter(.aggregated_cells>=10)

pseudo_bulk_filtered %>% as_tibble() %>% arrange(.aggregated_cells)

```


```{r}

library(tidyr)
library(dplyr)

# Pivot data: rows = samples, columns = features, values = assay data (X)
library(magrittr)
df <- pseudo_bulk_filtered
df$cell_types %<>% droplevels()
df$condition %<>% droplevels()

pca_matrix_full <- df %>%
  as_tibble() %>%
  select(.sample, .feature, X) %>%
  pivot_wider(names_from = .feature, values_from = X)

# Calculate the variance for each gene (column)
gene_variances <- apply(pca_matrix_full[,-1], 2, var)

# Take top 10% most variable
threshold <- quantile(gene_variances, 0.9, na.rm=T)

# Filter for highly variable genes
highly_variable_genes <- names(gene_variances[gene_variances >= threshold])
pca_matrix <- pca_matrix_full %>% select(all_of(c(".sample", highly_variable_genes)))

# Remove the `.sample` column and convert to matrix
pca_matrix_values <- pca_matrix %>%
  select(-.sample) %>%
  as.matrix()

# Check rownames
rownames(pca_matrix_values) <- pca_matrix$.sample

# Perform PCA
pca_result <- prcomp(pca_matrix_values, center = T, scale. = TRUE)

# View the PCA result
summary(pca_result)  # Check explained variance
pca_result$x         # Principal component scores for each sample

library(factoextra)
fviz_pca_ind(pca_result, label="none", habillage = df$cell_types, addEllipses = T, ellipse.level=0.95) + scale_color_brewer(palette="Dark2")

```

Now just neuts and just sham/naive

```{r}

df <- pseudo_bulk_filtered %>% filter(cell_types=="neutrophils") %>% filter(condition %in% c("Naive","Sham"))

df$cell_types %<>% droplevels()
df$condition %<>% droplevels()

pca_matrix_full <- df %>%
  as_tibble() %>%
  select(.sample, .feature, X) %>%
  pivot_wider(names_from = .feature, values_from = X)

# Calculate the variance for each gene (column)
gene_variances <- apply(pca_matrix_full[,-1], 2, var)

# Take top 10% most variable
threshold <- quantile(gene_variances, 0.9, na.rm=T)

# Filter for highly variable genes
highly_variable_genes <- names(gene_variances[gene_variances >= threshold])
pca_matrix <- pca_matrix_full %>% select(all_of(c(".sample", highly_variable_genes)))

```

Now add and integrate our stressed meninges neutrophils 
```{r}
# Load sce object
combi <- readRDS(paste0(here::here("res/"),"combi_clustered_leiden_annotated.RDS"))

load(file=here::here("res/amb_to_drop.R"))
sce <- combi[!rowData(combi)$ID %in% amb_to_drop,]
sce <- sce[,!sce$Annotation == "Erythrocytes"]
```

Extract neutrophils
```{r}
neuts <- sce[,sce$Annotation %in% "Neutrophils"]

# Make pseudobulk
pseudo_bulk_stress <- neuts %>%
  tidySingleCellExperiment::aggregate_cells(c(group, cond))

# Check cell numbers
pseudo_bulk_stress %>% colData %>% as_tibble() %>% arrange(.aggregated_cells)

# How many of their HVGs are in our dataset 
highly_variable_genes %in% rownames(pseudo_bulk_stress) %>% table()

# Perform PCA on the intersect of HVG in Kolabas and genes in stress
highly_variable_genes_intersect <- highly_variable_genes[highly_variable_genes %in% rownames(pseudo_bulk_stress)]

```

```{r}

# Subset stress to hvg intersect
pseudo_bulk_stress_hvg <- pseudo_bulk_stress %>%
  filter(.feature %in% highly_variable_genes_intersect) 

# Ensure reorder rows for both
pseudo_bulk_stress_hvg <- pseudo_bulk_stress_hvg[ highly_variable_genes_intersect, ]

# Reorder Kolabas to only HVG intersect
df <- df[highly_variable_genes_intersect, ]

# Check TRUE
identical(rownames(df), rownames(pseudo_bulk_stress_hvg))

```

Now run Kolabas PCA with HVG intersect
```{r}

pca_matrix <- df %>%
  as_tibble() %>%
  select(.sample, .feature, X) %>%
  pivot_wider(names_from = .feature, values_from = X)

# Remove the `.sample` column and convert to matrix
pca_matrix_values <- pca_matrix %>%
  select(-.sample) %>%
  as.matrix()

# Check rownames
rownames(pca_matrix_values) <- pca_matrix$.sample

# Perform PCA
pca_result <- prcomp(pca_matrix_values, center = T, scale. = TRUE)

# View the PCA result
summary(pca_result)  # Check explained variance
pca_result$x         # Principal component scores for each sample

```


Use same PCA model and apply to stress dataset 
```{r}
pca_matrix_stress <- pseudo_bulk_stress_hvg %>%
  as_tibble() %>%
  select(.sample, .feature, logcounts) %>%
  pivot_wider(names_from = .feature, values_from = logcounts)

# Remove the `.sample` column and convert to matrix
pca_matrix_stress_values <- pca_matrix_stress %>%
  select(-.sample) %>%
  as.matrix()

# Check rownames
rownames(pca_matrix_stress_values) <- pca_matrix_stress$.sample

# Scale the new sample using the PCA model parameters
new_samples_scaled <- scale(pca_matrix_stress_values, center = pca_result$center, scale = pca_result$scale)

# Project the scaled sample onto the PCA components by matrix multiplication
new_sample_pca <- new_samples_scaled %*% pca_result$rotation

# Combine original PCA scores with the new sample's PCA scores
pca_scores_combined <- as.data.frame(rbind(pca_result$x, new_sample_pca))
pca_scores_combined$sample <- rownames(pca_scores_combined)

pca_scores_combined$region <- str_split_i(rownames(pca_scores_combined),"___",2)

pca_scores_combined$condition <- str_split_i(rownames(pca_scores_combined),"___",3)

library(RColorBrewer)
# PC2 just represents condition (sham vs naive)
ggplot(pca_scores_combined, aes(x=PC2, y=PC3)) + geom_point(aes(colour=condition), size=4) + scale_color_brewer(palette = "Paired") + theme_light()

# So show PC1 and PC3
ggplot(pca_scores_combined, aes(x=PC1, y=PC3)) + geom_point(aes(colour=region), size=4) + scale_color_brewer(palette = "Paired") + theme_light()

```


```{r}

sessionInfo()

```
